---
title: "Framework: Peer Comparisons vs Internal Labor Stress Indicators"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
# Namespacing-only: no library() calls required
```

## Problem statement

Peer pay comparisons are often treated as if they are **market forces**. This framework separates:

- **Peer comparisons** (external reference points)
from
- **Internal labor stress indicators** (your actual market signal)

## Two different questions

### Q1: “What do peers pay?”
This is descriptive and externally referenced.

### Q2: “Are we experiencing labor scarcity (recruitment/retention stress)?”
This is diagnostic and internally evidenced.

A compensation decision should not substitute Q1 for Q2.

## The 2x2 matrix

- Peer pay position: lower / higher
- Internal stress: low / high

```{r}
#| label: peer-vs-stress-matrix
#| results: asis

matrix_df <- base::data.frame(
  `Peer pay position` = c("Lower than peers", "Lower than peers", "Higher than peers", "Higher than peers"),
  `Internal labor stress` = c("Low", "High", "Low", "High"),
  `Interpretation` = c(
    "No market evidence problem; peer gap alone is insufficient",
    "Possible market stress; target role-specific adjustment or redesign",
    "Overpay risk; verify capability needs and outcomes before sustaining",
    "Pay alone is not solving the problem; investigate non-pay constraints"
  ),
  `Primary actions` = c(
    "Hold; monitor; improve internal analytics",
    "Diagnose drivers; adjust pay where needed; fix job conditions",
    "Audit job scope and performance; ensure fiscal sustainability",
    "Investigate supervision/workload/tools; pipeline; training; culture"
  )
)

ft <- flextable::flextable(matrix_df)
ft <- flextable::autofit(ft)
ft
```

## Internal labor stress indicator set (recommended)

At minimum, report (by department and job family):

- Applicants per posting (and qualified applicants)
- Offer acceptance rate (and first-offer rejection rate)
- Time-to-fill
- Voluntary turnover and destination (other city / private / other)
- Early-tenure attrition (first 12–24 months)
- Overtime pressure (where relevant) as a symptom (not proof) of staffing strain

## Peer comparison as secondary evidence (proper role)

Peer benchmarks can be used to:

- Generate hypotheses (why are we different?)
- Identify *possible* competition sets (who is recruiting from whom?)
- Provide a sanity check on extreme outliers

They should **not** be used as a sufficient condition for raises.

## A defensible Council narrative (template)

A defensible write-up should include:

1. **Claim**: what labor problem exists (if any), stated narrowly.
2. **Evidence**: internal stress indicators over time, by role.
3. **Mechanism**: why pay is the binding constraint (vs non-pay factors).
4. **Intervention**: targeted pay adjustment or job redesign, with scope and cost.
5. **Evaluation**: what metrics will confirm improvement (applicants, acceptance, turnover).

## Minimal reporting package (what staff should bring)

- A one-page internal labor stress dashboard
- A role-specific recruiting funnel summary
- Exit destination analysis for voluntary departures
- Peer data appendix (clearly labeled as secondary)
