---
title: "Decision-Quality Rubric for Municipal Compensation Policy"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
# Namespacing-only: no library() calls required
```

## Purpose

This rubric evaluates the **quality of the decision process**, not merely the decision outcome. It is designed to prevent “peer studies” from replacing local evidence about:

- effectiveness
- developed capability
- actual labor-market constraints

## Scoring (0–4)

Each criterion is scored 0–4:

- 0 = absent / not addressed
- 1 = mentioned but unsupported
- 2 = partial evidence
- 3 = strong evidence and reasoning
- 4 = strong evidence, reasoning, and pre-committed evaluation plan

## Rubric table (flextable)

```{r}
#| label: decision-quality-rubric
#| results: asis

rubric_df <- base::data.frame(
  `Criterion` = c(
    "1. Problem definition",
    "2. Internal labor evidence",
    "3. Capability requirements",
    "4. Effectiveness / outcomes",
    "5. Peer comparison used appropriately",
    "6. Options considered (non-pay + pay)",
    "7. Fiscal sustainability",
    "8. Implementation details",
    "9. Evaluation plan",
    "10. Transparency & documentation"
  ),
  `What good looks like` = c(
    "Specific labor problem stated (which roles, what symptom, magnitude, timeframe)",
    "Applicant volume/quality, acceptance, time-to-fill, turnover, exit destinations by role",
    "Clear statement of required capabilities; scarcity indicators; job scope validated",
    "Performance evidence tied to role responsibilities (service levels, audit outcomes, risk control)",
    "Peer data is labeled as secondary; peer set selection rules disclosed; definitions aligned",
    "At least 3 alternatives: process/job redesign, pipeline/training, targeted pay adjustment",
    "Total compensation impact, compounding effects, and long-term obligations disclosed",
    "Who, when, which roles, pay bands, steps, and governance controls defined",
    "Pre-committed metrics + timeframe; what would count as success/failure",
    "All exhibits, assumptions, and definitions are accessible; limitations stated"
  ),
  `Score (0-4)` = base::rep("", 10),
  `Evidence / notes` = base::rep("", 10)
)

ft <- flextable::flextable(rubric_df)
ft <- flextable::autofit(ft)
ft
```

## How to use this rubric

1. Score each criterion independently.
2. Identify the **lowest-scoring criteria** (they usually reveal missing evidence).
3. Require staff to close the gaps *before* adopting a compensation change.
4. Attach the completed rubric to the agenda item as part of the public record.

## Red flags (process failures)

- “Peers pay more” presented as the primary justification without internal labor stress evidence.
- No data on offer acceptance, applicant quality, or exit destinations.
- Pay adjustments applied broadly without role-specific diagnosis.
- No plan to evaluate whether the change improved recruiting/retention/capability.

## What this rubric is not

- Not a claim that pay never matters.
- Not a substitute for management judgment.
- Not a political argument.

It is a tool for **decision integrity**: evidence, reasoning, and accountability.
